Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-08 Manuscript/manuscript.Rmd
edits

Biological replicate Hi-C data were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1) for the GM12878 and K562 cell lines. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for all resolutions and cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line.

Genomic annotation data in the form of cell type specific chromatin states (ChromHMM), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 1).

@@

To make the data amenable for model training, the genome was binned into equally sized intervals defined by the data resolution of the corresponding contact matrix. As an example, for domain data obtained from a 100 kb resolution contact matrix, the genome was binned into 100 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. A genomic bin was defined as a TAD boundary region ($Y = 1$) if it overlapped with at least one flanked TAD boundary, and a non-TAD boundary region ($Y = 0$) otherwise. Random forest models were used to train on Arrowhead-predicted TAD boundary regions. Training was performed separately for each cell line, data resolution, and chromosome.

Cell line-specific genomic annotations were used to built the predictors for of our random forest models. We created three types of predictors (Figure 1): **Overlap Count (OC)** - the total number of overlaps between a bin and genomic annotation regions;  **Overlap Percent (OP)** - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; **Log2 Distance** - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

The complete dataset used for modeling was composed of two components: the response vector **Y**, an indicator variable separating between TAD and non-TAD boundary regions, and the predictor-type specific data matrix, **X** = {$X_1$, $X_2$, ..., $X_p$}, where $p$ was the set of predictors. One, or all types of predictors were used as described. The data was split into a 7:3 ratio of training and testing sets.

### Addressing class imbalance

@@

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the set non-TAD boundary regions. CI of heavily imbalanced classes approaches 0, while CI for balanced classes approaches 0.50. To assess the effect of CI on predictive performance, we first used the data with **No Sampling**, that is, the majority and minority classes were used as is. We evaluated three re-sampling techniques. For **Random Under-Sampling (RUS)**, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For **Random Over-Sampling (ROS)**, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For **SMOTE** (Synthetic Minority Over-Sampling Technique), under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

### Assessment of model performance

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Ten-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. [??? Isn't it used for preventing overfitting? TBD] The number of trees that were aggregated for each random forest model was set to $n=500$.

The performance of each model was compared using Matthews Correlation Coefficient (MCC), defined as <!--The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. The MCC is defined as-->

@@

### Feature reduction

@@

## Using optimally built random forest classifiers for more precise TAD boundary prediction

Using optimally trained models, we established *predicted TAD boundary regions* as a collection of adjacent bases predicted as TAD boundary with the probability of 1. [??? Do not understand from here:] To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary regions*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. The midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a final *predicted TAD boundary*. [??? To here. REWRITE!]

@@

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** (A) Unique boundaries are denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e., 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a flanked TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Three types of associations between bins and genomic annotations (green shapes) used to build the feature space include _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)

@@

![**A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline.** Each cell line (GM12878, K562) specific model was composed of a combination of different data level characteristics including resolution (5 kb, 10 kb, 25 kb, 50 kb, 100 kb), predictor type (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)), and re-sampling technique (no re-sampling (None), random over-sampling (ROS), random under-sampling (RUS), synthetic minority over-sampling technique (SMOTE)). The three types of genomic annotations, histone modification (HM), chromating state segmentations (ChromHMM), transcription factor binding sites (TFBS), were used as predictors.](Figures/figure3_ensemble_framework_V3.png)

@@

![**Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal MCCs across resolution.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v3_bars.png)

@@

![**Assessing top ranked transcription factors in TAD boundary prediction.** (A) Barplots comparing performances of models using specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three groups (Combined; purple). (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes. Error bars represent standard deviation from the mean MCC across each chromosome.](Figures/annot_specific_rfe_varimps_heatmap.png)

@@

![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17390000-18640000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.  Called TADs are outlined in blue.  The blue vertical lines represent a called TAD boundary identified by the ARROWHEAD algorithm. The black line represents the predicted probability of each base pair being a TAD boundary. Regions of adjacent base pairs with probability of 1 are represented by the green regions along the x-axis. The green vertical lines represent the subsequent base pair predicted to be a TAD boundary using medoid partitioning. (B) Enrichment heatmaps illustrate the average signal strength of ChIP-seq peak regions defined by SMC3, ZNF143, CTCF, and RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.](Figures/visualizing_boundaries_example_with_enriched_heatmaps.png)

>>>>

Biological replicate Hi-C data were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1) for the GM12878 and K562 cell lines. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from Hi-C data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the ARROWHEAD algorithm [@durand2016juicebox]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for all resolutions and cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line.

Genomic annotation data in the form of cell type-specific cis-regulatory elements (CSS), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 2).

@@

To make the data amenable for model training, the genome was first binned into equally sized intervals defined by the data resolution (*r*) of the corresponding contact matrix. The binning was initiated at the chromosome-specific genomic coordinate marking half of the resolution, and continued in congruent intervals until the end of the chromosome (*mod r* + *r/2*). This was to ensure that called (ground truth) TAD boundaries would be positioned in the center of genomic bins. Genomic bins were then defined as TAD boundary regions (*Y = 1*) if they contained a TAD boundary, and non-TAD boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for random forest classification.

Cell type-specific genomic annotations were used to build the predictor space for of our random forest models. We created three types of predictors: *Overlap Count (OC)* - the total number of overlaps between a bin and genomic annotation regions;  *Overlap Percent (OP)* - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; *Log2 Distance* - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

The chromosome-specific dataset used for modeling was composed of two components: the response vector **Y**, an indicator variable separating between TAD and non-TAD boundary regions, and the predictor-type specific data matrix, **X** = {$X_1$, $X_2$, ..., $X_p$}, where $p$ was the set of genomic annotations. <!--One, or all types of predictors were used as described. The data was split into a 7:3 ratio of training and testing sets.-->

## Addressing class imbalance

@@

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the set non-TAD boundary regions. CI of heavily imbalanced classes approaches 0, while CI for balanced classes approaches 0.50. To assess the effect of CI on predictive performance, we first used the data with *No Sampling*, that is, the majority and minority classes were used as is. We evaluated three re-sampling techniques. For *Random Under-Sampling (RUS)*, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For *Random Over-Sampling (ROS)*, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For *SMOTE* (Synthetic Minority Over-Sampling Technique), under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

## Assessment of model performance

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Matthews Correlation Coefficient (MCC), defined as <!--The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. The MCC is defined as-->

@@

## Feature reduction

@@

## Precise TAD boundary prediction at bp resolution

Next, we aimed to investigate whether we could alleviate resolution limitations that plague conventional TAD calling algorithms. To do so, we extended our optimally built TAD prediction models to obtain predictions at base pair (bp) resolution. Naturally, when dealing with such fine resolution, there will exist collections of spatially close base pairs with similar predictor profiles (i.e. similar distances), and will thus yield similar predictions. To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both hierarchical clustering and partitioning techniques to identify more precise TAD boundaries throughout the linear genome.

We first established a *predicted TAD boundary area* (PTBA) as a collection of adjacent base pairs that have been identified as being a predicted TAD boundary (using the optimal probability threshold identified by the combination of TPR and FPR closest to the point (0,1) on the ROC curve). We then performed hierarchical clustering on the pairwise distances (in base pairs) between the midpoints of each PTBA using complete linkage clustering. To identify the optimal number of clusters, *k*, we calculated the mean silhouette score over all possible sizes of clusters *m*:

$$\bar{s}_{m} = \dfrac{\sum^{m}_{i} s_{i}}{m},$$

where $s_{i}$ is the cluster-specific silhouette score defined as,

$$s_{i} = \dfrac{b_{i} - a_{i}}{max[a_{i},b_{i}]}.$$

Here, $a_{i}$ is the mean distance between each cluster entry and all other observations in the same cluster, and $b_{i}$ is the mean distance between entries within the same cluster. Therefore number of clusters, *k*, that maximized the mean silhouette score, $\bar{s}_{m}$ were used to identify groupings of PTBA, which we defined as *predicted TAD boundary regions* (PTBR).

To precisely identify the PTBA most representative of each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The midpoint of the corresponding cluster medoid, in this case a particular PTBA, will be defined as a *predicted TAD boundary point* (PTBP). In the event that a particular PTBA has an even number of base pairs, then ceiling function was applied to extract the right most base pair. Thus, the hierarchical structure of our defined predicted TAD boundary objects were as follows: PTBR > PTBA > PTBP. That is, each PTBR is a collection of PTBAs, which are themselves a collection of adjacent base pairs. The uniquely identified PTBP was then the midpoint of the most representative PTBA within a PTBR.

<!--
Using optimally trained models, we established *predicted TAD boundary regions* as a collection of adjacent bases predicted as TAD boundary with the probability of 1. [??? Do not understand from here:] To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary regions*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. The midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a final *predicted TAD boundary*. [??? To here. REWRITE!]
-->

## Comparing Called vs Predicted TAD boundaries

To evaluate if our predicted TAD boundaries are biologically significant we will asssess the signal strength obtained from key ChIP-Seq peak elements known to be enriched at TAD boundaries including CTCF, RAD21, SMC3, and ZNF143. To do so we will used enrich heatmaps and Student’s t-tests to compare distances from selected ChiP-seq data to our predicted TAD boundaries compared to called TAD boundaries (ARROWHEAD).

Furthermore, we will assess the conservation of our predicted TAD boundaries within cell types and between resolutions, as well as between cell types and within resolutions and compare with ARROWHEAD called TAD boundaries. To do so we will evaluate Jaccard Indices. Jaccard is a metric used to measure overlap between sets (genomic regions). Thus we can use it as measure of overlap between TAD boundaries. Given two sets of TAD boundaries (either between resolution or cell type) A and B, we define the Jaccard Index as,

$$J_{(A,B)} = \dfrac{A \cap B}{A \cup B}.$$
To give the sets of boundaries A and B dimensionality to enable us to calculate overlap, we flank each boundary point by a flanking factor, *f*=5000.

@@

<!--
![**Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** (A) Unique boundaries are denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e., 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a flanked TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Three types of associations between bins and genomic annotations (green shapes) used to build the feature space include _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)
-->

@@

<!--
![**A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline.** Each cell line (GM12878, K562) specific model was composed of a combination of different data level characteristics including resolution (5 kb, 10 kb, 25 kb, 50 kb, 100 kb), predictor type (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)), and re-sampling technique (no re-sampling (None), random over-sampling (ROS), random under-sampling (RUS), synthetic minority over-sampling technique (SMOTE)). The three types of genomic annotations, histone modification (HM), chromating state segmentations (ChromHMM), transcription factor binding sites (TFBS), were used as predictors.](Figures/figure3_ensemble_framework_V3.png)
-->

@@

<!--
![**Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal MCCs across resolution.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v3_bars.png)
-->

@@

<!--
![**Assessing top ranked transcription factors in TAD boundary prediction.** (A) Barplots comparing performances of models using specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three groups (Combined; purple). (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes. Error bars represent standard deviation from the mean MCC across each chromosome.](Figures/annot_specific_rfe_varimps_heatmap.png)
-->

@@

<!--
![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17390000-18640000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.  Called TADs are outlined in blue.  The blue vertical lines represent a called TAD boundary identified by the ARROWHEAD algorithm. The black line represents the predicted probability of each base pair being a TAD boundary. Regions of adjacent base pairs with probability of 1 are represented by the green regions along the x-axis. The green vertical lines represent the subsequent base pair predicted to be a TAD boundary using medoid partitioning. (B) Enrichment heatmaps illustrate the average signal strength of ChIP-seq peak regions defined by SMC3, ZNF143, CTCF, and RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.](Figures/visualizing_boundaries_example_with_enriched_heatmaps.png) 
-->
